### 2018-04-13 paper/appendix.tex
pass over most of correctness proof

\section{Proof of Correctness}
\label{app:corr}
%
In this section, we give a proof that \sys is a correct solver for
recursion-free CHC systems.
%
We first establish lemmas for the correctness of each procedure used
by \sys, namely \textsc{Collapse} (\autoref{lem:expansion-sound} and
\autoref{lem:expansion-complete}), $\expand$
(\autoref{lem:expand-corr}), and $\solvecdd$
(\autoref{lem:vc}, \autoref{lem:cdd-soln-sound}, and
\autoref{lem:cdd-soln-complete}).
%
We combine the lemmas to prove correctness of \sys
(\autoref{thm:corr}).

% lem: result connecting expansions to constructing solutions
For recursion-free CHC systems $S$ and $S'$, if $S'$ is
an expansion of $S$, then the result of collapsing a
solution of $S'$ is a solution of $S$.
%
\begin{lem}
  \label{lem:expansion-sound}
  For recursion-free CHC system $S'$ and $S$,
  $\sigma'$ is a solution of $S'$, and $\eta : \predof{S}' \to
  \predof{S}$ a correspondence from $\predof{S}'$ to $\predof{S}$,
  $\collapse{\eta}{\sigma'}$ is a solution of $S$.
\end{lem}
%
\begin{proof}
  % introduce 
  Let $\mathcal{B} \in \bodies{\mathcal{R}}$, $\varphi \in \formulas$,
  and $R \in \mathcal{R}$ be such that $((\mathcal{B}, \varphi),
  R(\vars{ \mathcal{S} }(R))) \in \mathcal{S}$.
  %
  For each predicate $P' \in \predof{S}'$ such that $\eta(P') = P$, there is
  some clause $C' \in S'$ such that $P' \in \bodyof{C'}$ by the fact that $S'$ is an expansion of
  $S$ and the definition of an expansion.
  %
  Let preicate $Q' \in \predof{S}'$ such that $Q'$ is the head of clause $C'$.
  % 
  $\sigma'[P'] \land \consof{C'} \entails \sigma'[Q']$ by
  the fact that $\sigma'$ is a solution of $S'$.
  %
  Therefore,
  %
  \[ \collapse{ \eta }{ \sigma' }[P] \land \consof{C'} \entails %
  \left( \bigland_{ \substack{Q' \in \predof{S}' \\ \eta(Q') = Q} }
  \sigma(Q') = %
  \collapse{\eta}{\sigma'}[Q'] \right)
  \]
  % 
  Therefore, $\collapse{\eta}{\sigma'}$ is a solution of $P$.
  %
  Therefore, $\collapse{\eta}{\sigma'}$ is a solution of
  $S$.
\end{proof}

% completeness of expansions:
For a recursion-free CHC system $S$ is solvalbe, each expansion of $S$
is solvable.
\begin{lem}
  \label{lem:expansion-complete}
  For a recursion-free CHC system $S$ is solvable, and
  $S'$ is
  an expansion of $S$, $S'$ is solvable.
\end{lem}
%
\begin{proof}
  Let $\sigma : P \to \formulas$ be a solution of
  $S$, and let $\eta : \predof{S}' \to \predof{S}$ be a
  correspondence from $S'$ to $S$.
  %
  Let $\sigma' : \predof{S}' \to \formulas$ be such that for each $P'
  \in \predof{S}'$, $\sigma'(P') = \sigma(\eta(P'))$.
  %
  Then $\sigma'$ is a solution of $S'$.
\end{proof}

% CDD systems:
$\expand$ always returns a CDD expansion of its input.
%
\begin{lem}
  \label{lem:expand-corr}
  For a recursion-free CHC system $S$ and $S'$, and $\eta : \predof{S'} \to \predof{S}$ such that
  $(S', \eta) = \expand(S)$, $\eta$
  is a correspondence from $S'$ to $S$, $S'$ is an CDD system and an expansion
  of $S$.
\end{lem}
%
\begin{proof}
  Proof by induction on the evaluation of $\expand$ on
  an arbitrary recursion-free CHC system $S$.
  % inductive fact:
  The inductive fact is that in each step of evaluation \expandaux,
  $\corr$ is a correspondence from argument
  $S'$ to $S$.
  % base case:
  For the base case, \expandaux is called initially on $S$,
  by \autoref{alg:expand}.
  % 
  $\corr$ is a correspondence from $S$ to
  itself, by the definition of $\corr$
  (\autoref{app:cons-cdd}).

  % inductive case:
  For the inductive case, for $C \in S$ and $P \in \predof{S}$, $\expandaux$ calls itself on 
  $\copyrel(S, C, P)$.
  %
  For each recursion-free CHC system $S'$ that generated by $\copyrel(S, C, P)$,
  $\corr$ is a correspondence from $S'$ to
  $S$ by definition of $\copyrel$
  (\autoref{app:cons-cdd}).
  %
  By this fact and the inductive hypothesis, $\corr$ is
  a correspondence from $\copyrel(S',
  C, P)$ to $S$.

@@

% lem: VC's characterize system sat:
For each recursion-free CHC system $S$, $S$ has a solution if and
only if all interpolants queries return interpolants.
%
\begin{lem}
  \label{lem:vc}
  Given a recursion-free CHC system $S$ that is CDD and solvable, for all predicates $P \in \predof{S}$,
   $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns a interpolatn $I$.
\end{lem}
%
\begin{proof}
  %
  Assume that $S$ has solution $\sigma$ and there are
  some predicates $P \in \predof{S}$ such that $\solveitp(\prectr(\cc{P},\sigma'),\postctr(\cc{P},\sigma'))$
  returns $SAT$, which means there is a model $m$ for the conjuntion of $\prectr(\cc{P},\sigma')$
  and $\postctr(\cc{P},\sigma')$.
  %
  But $\postctr(\cc{P},\sigma) = \false$, by the definition of
  a solution of a CHC system.
  %
  Therefore, there can be no such model $m$.
  %
  Therefore, $\solveitp(\prectr(\cc{P},\sigma'),\postctr(\cc{P},\sigma'))$ always returns interpolant.
\end{proof}

\begin{lem}
  \label{lem:solve-aux}
  Given a recursion-free CHC system $S$ that is CDD, if for all predciates $P \in \predof{S}$, 
  $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns a interpolatn $I$, 
  then $\solvecdd(S)$ is a solution of $S$.
\end{lem}
%
\begin{proof}
  % overview of structure:
  Proof by induction on the $\solvecdd(S)$ calles $\solveitp$ on all predicates $P \in \predof{S}$
  in a topolgical order.
  %
  The inductive fact is that at each step of calling $\solveitp$, $\sigma$ is a
  partial solution of $S$.
  % base case:
  For the base case, $\predof{S} = \varnothing$,
  %
  Therefore, $\sigma$ is a solution of $S$.

  % inductive case:
  For the inductive case, in each step, $\solvecdd$ calls $\solveitp$ on
  a predicate $P \in \predof{S}$ with partial solution $\sigma$, which
  contains all solutions of predciates $P \in \depsof{S}$.
  %
  Based on the definition of an interpolant(\autoref{defn:itps}), $\sigma$
  is a partial solution of $S$, then $\prectr(\cc{P},\sigma')$ and $\postctr(\cc{P},\sigma))$
  are inconsistent.
  %
  The interpolant $I$ returns by $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ entails
  by all clause that head is $P$ and $\bodyof{P}$ is substituted by partial solution $\sigma$.
  %
  $I$ also is inconsistment with all constrains shows later that support the query clause.
  %
  Then $\solvecdd$ updates $\sigma$ by binding $P$ to $I$, which makes the new $\sigma$ also a partial
  solution of $S$.

  %
  Base case combined with the inductive case, imply that \solvecdd
  returns a solution of $S$.
\end{proof}

% SolveCDD: establish soundness
$\solvecdd$ is a
correct CHC solver for CDD systems $S$.
%
\begin{lem}
  \label{lem:cdd-soln-sound}
  For a given CDD system $S$, and $\sigma = \solvecdd(S)$, %
  $\sigma$ is a solution of $\mathcal{S}$.
\end{lem}
%
\begin{proof}
  %
  $\solvecdd{S}$ returns $\sigma$ implies that for all predicates 
  $P \in \predof{S}$, $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns a interpolatn $I$, 
  by the definition of $\solvecdd$ \autoref{alg:solve-cdd}. 
  %
  Therefore, \autoref{lem:solve-aux} implies that \solvecdd
  returns a complete solution of $S$.
\end{proof}

>>>>

\section{Proof of Correctness}
\label{app:corr}
%
In this section, we prove that \sys is correct when applied to
recursion-free CHC systems.
%
We first establish lemmas for the correctness of each procedure used
by \sys, namely \textsc{Collapse} (\autoref{lem:expansion-sound} and
\autoref{lem:expansion-complete}), $\expand$
(\autoref{lem:expand-corr}), and $\solvecdd$
(\autoref{lem:vc}, \autoref{lem:cdd-soln-sound}, and
\autoref{lem:cdd-soln-complete}).
%
We combine the lemmas to prove \sys is correct (\autoref{thm:corr}).

% lem: result connecting expansions to constructing solutions
For two recursion-free CHC systems $S$ and $S'$, if $S'$ is
an expansion of $S$, then the result of collapsing a
solution of $S'$ is a solution of $S$.
%
\begin{lem}
  \label{lem:expansion-sound}
  For two recursion-free CHC system $S'$ and $S$ such that $\sigma'$
  is a solution of $S'$ and $\eta$ is a correspondence from
  $S'$ to $S$, $\collapse{\eta}{\sigma'}$ is a
  solution of $S$.
\end{lem}
%
\begin{proof}
  % introduce 
  %
  For each predicate $P' \in \predof{S}'$ such that $\eta(P') = P$,
  there must exist some clause $C' \in S'$ such that $P' \in \bodyof{C'}$
  because $S'$ is an expansion of $S$.
  %
  Let predicate $Q' \in \predof{S}'$ be the head of $C'$.
  % 
  $\sigma'[P'] \land \consof{C'} \entails \sigma'[Q']$ by
  the fact that $\sigma'$ is a solution of $S'$.
  %
  Therefore,
  %
  \[ \collapse{ \eta }{ \sigma' }[P] \land \consof{C'} \entails %
  \left( \bigland_{ \substack{Q' \in \predof{S}' \\ \eta(Q') = Q} }
  \sigma(Q') = %
  \collapse{\eta}{\sigma'}[Q'] \right)
  \]
  % 
  Therefore, $\collapse{\eta}{\sigma'}$ has a solution for $P$.
  %
  Since $\collapse{\eta}{\sigma'}$ has a solution for each predicate
  in $S$, $\collapse{\eta}{\sigma'}$ is a solution of $S$.
\end{proof}

Every expansion of a solvable recursion-free CHC system is also
solvable.
% completeness of expansions:
\begin{lem}
  \label{lem:expansion-complete}
  If a recursion-free CHC system $S$ is solvable and
  $S'$ is an expansion of $S$, then $S'$ is solvable.
\end{lem}
%
\begin{proof}
  Let $\sigma$ be a solution of $S$, and let $\eta$ be a
  correspondence from $S'$ to $S$.
  %
  Let $\sigma'$ be such that for each $P' \in \predof{S}'$,
  $\sigma'(P') = \sigma(\eta(P'))$.
  %
  Then $\sigma'$ is a solution of $S'$.
\end{proof}

% CDD systems:
$\expand$ always returns a CDD expansion of its input.
%
\begin{lem}
  \label{lem:expand-corr}
  For two recursion-free CHC systems $S$ and $S'$ and a correspondence
  from $S'$ to $S$, $\eta$, such that $(S', \eta) = \expand(S)$,
  $S'$ is an CDD system and an expansion of $S$.
\end{lem}
%
\begin{proof}
  By induction over the evaluation of $\expand$ on
  an arbitrary recursion-free CHC system $S$.
  % inductive fact:
  The inductive fact is that for each evaluation step, \expandaux,
  $\corr$ is a correspondence from argument
  $S'$ to $S$.
  % base case:
  In the base case, \expandaux is called initially on $S$,
  by \autoref{alg:expand}.
  %
  $\corr$ is a correspondence from $S$ to itself, by the definition of
  $\corr$
  (\autoref{app:cons-cdd}).

  % inductive case:
  In the inductive case,
  $\expandaux$ constructs an argument
  $\copyrel(S, C, P)$
  where $C$ is a clause and $P$ is a predicate in $S$.
  $\expandaux$ recusively invokes itself with this argument.
  %
  For each recursion-free CHC system $S'$ generated by $\copyrel(S, C, P)$,
  $\corr$ is a correspondence from $S'$ to
  $S$ by definition of $\copyrel$
  (\autoref{app:cons-cdd}).
  %
  By this fact and the inductive hypothesis, $\corr$ is
  a correspondence from $\copyrel(S',
  C, P)$ to $S$.

@@

% lem: VC's characterize system sat:
For each recursion-free CHC system $S$, $S$ has a solution if and only
if all interpolants queries return interpolants.
%
\begin{lem}
  \label{lem:vc}
  Given a recursion-free CHC system $S$ that is CDD and solvable, for
  all predicates $P \in \predof{S}$,
  $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns
  an interpolant $I$.
\end{lem}
%
\begin{proof}
  %
  Assume that $S$ has a solution $\sigma$ and there are some
  predicates $P \in \predof{S}$ such that
  $\solveitp(\prectr(\cc{P},\sigma'),\postctr(\cc{P},\sigma'))$
  returns $SAT$. This means there must be a model $m$ for the
  conjunction of $\prectr(\cc{P},\sigma')$ and
  $\postctr(\cc{P},\sigma')$.
  %
  But $\postctr(\cc{P},\sigma) = \false$, by the definition of
  a solution of a CHC system.
  %
  Therefore, there can be no such model $m$.
  %
  Therefore,
  $\solveitp(\prectr(\cc{P},\sigma'),\postctr(\cc{P},\sigma'))$ always
  returns interpolant.
\end{proof}

\begin{lem}
  \label{lem:solve-aux}
  Given a recursion-free CHC system $S$ that is CDD, if for all
  predicates $P \in \predof{S}$,
  $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns
  an interpolant $I$, then $\solvecdd(S)$ is a solution of $S$.
\end{lem}
%
\begin{proof}
  % overview of structure:
  By induction on the $\solvecdd(S)$ calls to $\solveitp$ over all
  predicates $P \in \predof{S}$
  in topological order.
  %
  The inductive fact is that after each call to $\solveitp$, $\sigma$
  is a partial solution of $S$.
  % base case:
  In the base case, $\predof{S} = \varnothing$,
  %
  Therefore, $\sigma$ is a solution of $S$.

  % inductive case:
  In the inductive case, $\solvecdd$ calls $\solveitp$ on a predicate
  $P \in \predof{S}$ with partial solution $\sigma$.  Due to the
  topological ordering, $\sigma$ contains interpretations for each
  predicate $P \in \depsof{S}$.
  %
  Based on the definition of an interpolant(\autoref{defn:itps}),
  $\prectr(\cc{P},\sigma)$ and $\postctr(\cc{P},\sigma))$ are
  inconsistent.
  %
  The interpolant $I$ returned by
  $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ is
  entailed by each clause $C$ where $P$ is the head where the
  predicates in $\bodyof{C}$ are substituted by their interpretations
  in$\sigma$.
  %
  $I$ is also inconsistment with all constraints that appear after $P$
  that support the query clause.
  %
  Therefore, when $\solvecdd$ updates $\sigma$ by binding $P$ to $I$
  the result is a partial solution of $S$.
\end{proof}

% SolveCDD: establish soundness
$\solvecdd$ is correct for CDD systems.
%
\begin{lem}
  \label{lem:cdd-soln-sound}
  For a given CDD system $S$, and $\sigma = \solvecdd(S)$, %
  $\sigma$ is a solution of $S$.
\end{lem}
%
\begin{proof}
  %
  The fact that $\solvecdd(S)$ returns $\sigma$ implies that for each
  predicate $P \in \predof{S}$,
  $\solveitp(\prectr(\cc{P},\sigma),\postctr(\cc{P},\sigma))$ returns
  a valid interpolant (\autoref{alg:solve-cdd}).
  %
  Therefore, \autoref{lem:solve-aux} implies that \solvecdd~returns a
  complete solution of $S$.
\end{proof}

### 2018-04-13 paper/evaluation.tex
pass over all except proof of correctness
pass over most of correctness proof

\begin{figure}[t]
  \centering
  \begin{floatrow}[2]
    \ffigbox[.48\textwidth] %
    {\caption{Solving times of \sys vs. \duality.
        %
        The $x$ and $y$ axes range over the solving times in seconds
        of \sys and \duality, respectively.
        %
        Each point depicts the performance of a benchmark.
        %
        The line $y = x$ is shown in red.} %
      \label{fig:complete-data} }
    { \includegraphics[width=\linewidth]{fig/complete.png} }
    %
    \ffigbox[.48\textwidth] %
    { \caption{Times of \sys and \duality vs. system size.
        %
        The $x$-axis ranges over the size of a given system, and the
        $y$-axis ranges over solvers' times.
        %
        Measurements of \sys and \duality are shown in blue and
        red, respectively. } %
      \label{fig:size} }
     { \includegraphics[width=\linewidth]{fig/size.png} }
  \end{floatrow}
\end{figure}
% introduce figure, summarize data not included:
The results of our evaluation are in
\autoref{fig:complete-data}.
%
Of the 4,040 benchmarks on which both solvers took a short amount
of time---less than five seconds---\sys solved the benchmarks in an
average of 0.51 seconds and \duality solved them in an average of 0.42
seconds.
%
\autoref{fig:complete-data} contains data for benchmarks which took
longer than five second for both systems to solve.
%
Out of these 269 benchmarks, \sys solved 185 in less time
than \duality, and solved 159 in less than half the time of \duality.
%
\duality solved 84 in less time than \sys, and solved 53 in less than
half the time of \sys.
%
Of the 762 benchmarks on which \sys timed out, \duality solved or
found a counterexample to 185.
%
Of the 1,145 benchmarks on which \duality timed out, \sys solved or
found a counterexample to 470.

@@

%result is good
The results indicate that \textbf{(1)} on a significant number of
different verification problems, \sys can perform significantly better than
\duality, but
%
\textbf{(2)} there are some cases in which the strengths of each
algorithm yield better results.
%
We collected the differences between sizes of a given system and its
minimal CDD expansion generated by \sys and found that they were
independent of \sys's performance compared to \duality.
%
Thus, while \duality may in the worst case enumerate exponentially
many derivation trees, it appears to enumerate far fewer
than the worst-case bound in some cases, causing it to perform better
than \sys.
%
Our results indicate that a third approach that combines the strengths
of both \duality and \sys, perhaps by lazily unwinding a given
recursion free system into a series of CDD systems instead of
derivation trees, could yield further improvements.

>>>>

\begin{figure}[t]
  \centering
  \begin{floatrow}[2]
    \ffigbox[.48\textwidth] %
    {\caption{Solving times of \sys vs. \duality.
        %
        The $x$ and $y$ axes range over the solving times in seconds
        of \sys and \duality, respectively.
        %
        Each point depicts the performance of a benchmark.
        %
        The line $y = x$ is shown in red.} %
      \label{fig:complete-data} }
    { \includegraphics[width=\linewidth]{fig/complete.png} }
    %
    \ffigbox[.48\textwidth] %
    { \caption{Times of \sys and \duality vs. system size.
        %
        The $x$-axis ranges over the size of a given system, and the
        $y$-axis ranges over solvers' times.
        %
        Measurements of \sys and \duality are shown in blue and
        red, respectively. } %
      \label{fig:size} }
     { \includegraphics[width=\linewidth]{fig/size.png} }
  \end{floatrow}
\end{figure}
% introduce figure, summarize data not included:
The results of our evaluation are shown in
\autoref{fig:complete-data}.
%
Of the 4,040 benchmarks on which both solvers took a short amount
of time---less than five seconds---\sys solved the benchmarks in an
average of 0.51 seconds and \duality solved them in an average of 0.42
seconds.
%
\autoref{fig:complete-data} contains data for benchmarks which took
longer than five second for both systems to solve.
%
Out of these 269 benchmarks, \sys solved 185 in less time
than \duality, and solved 159 in less than half the time of \duality.
%
\duality solved 84 in less time than \sys, and solved 53 in less than
half the time of \sys.
%
Of the 762 benchmarks on which \sys timed out, \duality solved or
found a counterexample to 185.
%
Of the 1,145 benchmarks on which \duality timed out, \sys solved or
found a counterexample to 470.

@@

%result is good
The results indicate that \textbf{(1)} on a significant number of
different verification problems, \sys can perform significantly better than
\duality, but
%
\textbf{(2)} there are some cases in which the strengths of each
algorithm yield better results.
%
We collected the differences between sizes of a given system and its
minimal CDD expansion generated by \sys and found that they were
independent of \sys's performance compared to \duality.
%
Thus, while \duality may in the worst case enumerate exponentially
many derivation trees, it appears to enumerate far fewer
than the worst-case bound in some cases, causing it to perform better
than \sys.
%
Our results indicate that a third approach that combines the strengths
of both \duality and \sys, perhaps by lazily unwinding a given
recursion-free system into a series of CDD systems instead of
derivation trees, could yield further improvements.

### 2018-04-13 paper/introduction.tex
pass over most of correctness proof

% contribution of this paper: CDD systems:
The first contribution of this paper is the introduction of a novel
class of directly solvable systems that we refer to as
\emph{Clause-Dependence Disjoint} (CDD).
%
The class consists of all recursion-free CHC systems such that each
clause in the system has at most one dependency on the same relational
predicate.
%
The key characteristic of this class is that when an arbitrary
recursion-free system is reduced to a CDD system and to a system from
a different directly solvable class, the CDD system is frequently the
smaller of the two.
%
Therefore, solving recursion free systems by reducing them to CDD form
is often less computationally expensive than reducing them to a
system in a different class.

>>>>

% contribution of this paper: CDD systems:
The first contribution of this paper is the introduction of a novel
class of directly solvable systems that we refer to as
\emph{Clause-Dependence Disjoint} (CDD).
%
The class consists of all recursion-free CHC systems such that each
clause in the system has at most one dependency on the same relational
predicate.
%
The key characteristic of this class is that when an arbitrary
recursion-free system is reduced to a CDD system and to a system from
a different directly solvable class, the CDD system is frequently the
smaller of the two.
%
Therefore, solving recursion-free systems by reducing them to CDD form
is often less computationally expensive than reducing them to a
system in a different class.
